Venn diagrams may indicate erroneous statistical reasoning in transcriptomics.

### 2022-03-08 mk5/manuscript.rmd
version 5

---
title: "Venn diagrams as indicators of erroneous statistical reasoning in transcriptomics."
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
outputoff:
  html_document
output:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

## Titles

???

@@

A common application of differential expression analysis is to
find genes that are differentially expressed upon treatment in only one
out of several groups of samples. One of the approaches is to test 
for significant difference in expression between treatment and control separately
in the two groups, and then select genes that show statistical
significance in one group only. This approach is then often combined with
a gene set enrichment analysis to find pathways and gene sets
regulated by treatment in only this group.

Here we show that this procedure is statistically incorrect and that the interaction
between treatment and group should be tested instead. Moreover, we show
that gene set enrichment analysis applied to such incorrectly defined genes
group-specific genes only may result in misleading artifacts. Genes which
show significant difference upon treatment in one, but not another group
are enriched in gene sets associated with the treatment in general.  Thus,
the results appear related to the problem at hand, but do not reflect the
group-specific effect of a treatment. A literature search revealed that
more than a quarter of papers which used a Venn diagram to illustrate the
results of separate differential analysis  have also applied this incorrect
reasoning.

@@

Experimental designs for transcriptomic analyses commonly include more than one factor. Often, the
question asked is whether there is a difference between groups (first
factor) with respect to reaction to a particular treatment (second
factor). For example, we may ask whether there are differentially
expressed genes (DEGs) which are specific to a particular group of
patients, e.g. interferon response elicited by a virus in one group, but
not regulated in another group of patients. In other words, we ask whether the
difference between control (healthy subjects) and treatment (infected
patients) is different between two groups of individuals. This "difference
of differences" is known in statistics as an interaction
[@blalock1965theory]. To find out whether it is statistically significant,
an appriopriate statistical test for interaction should be employed.

However, another approach is widely spread [@nieuwenhuis2011erroneous].
Instead of testing the interaction, the effect of the treatment is tested
separately in both groups. Next, a difference between groups is inferred if the difference
is significant in one comparison, but not significant in the other. This
approach is not correct from statistical point of view, as "the difference
between significant and not significant is not itself statistically
significant" [@gelman2006difference]. For example, the p-value in the first
comparison may be 0.009, and in the other comparison 0.011, thus at an
alpha level of 0.01 the difference will be statistically significant in the
first, but not significant in the other comparison.

In transcriptomics, statistical tests are performed for thousands of
genes. As in the general case, the inference of differences between the
groups should correctly be done by testing the significance of interaction
between the group and the treatment. However, often the differences between
treatment and control are tested in the two groups separately, and the
results are then compared. This can be visualized using a Venn diagram (VD,
Fig. 1) showing the overall number of DEGs significant in both comparisons
(the intersection in the VD) or significant in only one comparison (the
remaining two fields on a VD).  The genes which are significant in only one
comparison may then be incorrectly defined as specifically to the corresponding
group. Following this, gene set enrichment analysis may be applied to
these genes in an attempt to test which pathways are specific to one, but
not the other group.

In this paper, we show that under reasonable assumptions this approach may
result in apparent enrichments even if there are no real differences
between the groups. To this end, we randomly split a cohort into two
groups, compared the treatment (viral infection) with controls in each of
the groups separately and then applied gene set enrichment to the sets of
genes significantly different in one, but not the other group.
Moreover, we show that the resulting gene set enrichments correspond to the
differential expression between treatment and control. Thus, the enriched
terms are relevant to the studied project, yet while they do reflect real
processes linked to treatment, they are not demonstrative of differences
between the study groups. Finally, we use literature search to show that
this incorrect approach to study group-specific treatment effects is not
uncommon. In fact, while VDs are a useful visualisation tool also in
transcriptomics, in more than quarter of the papers where VDs were used,
group-specific genes were defined as significant in one, but not other
groups, and in 19% a gene set enrichment was performed.

@@

Consider two group of patients, G1 and G2 (Tab. 1). Each group contains 
`r n_stratum * 4`
individuals. In both groups, there is an equal number of healthy
individuals (labeled "Ctrl" on figures below) 
or patients infected with Sars-Cov-2 (labeled "SC2"). Our aim is to understand
the differences between G1 and G2 in the response to infection. For
example, we ask which genes or pathways are specifically upregulated by SC2
infection in G1 as compared to G2, and vice versa. In the following, we
used the data set GSE156063 [@mick2020upper] in two approaches (an
incorrect and the correct one) which arrive at opposite conclusions.

@@

First, we have performed differential gene expression analysis for each of
the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
DEGs
by using an false discovery rate (FDR) threshold of 
`r padj_thr`
and absolute log~2~ fold change (LFC) threshold of
`r lfc_thr`. There were
`r sum(res$g1$DEG)` 
DEGs in the G1 group, and
`r sum(res$g2$DEG)`
in the G2 group. In total,
`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
DEGs were common for G1 and G2, 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
DEGs were significant in G1 only ("specific" for G1), and 
`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
were significant in G2 only (see Fig. 1, panel *A*).
A naive interpretation of these results implies that there is a substantial
difference between these two groups of individuals, as evidenced by a small
overlap in commonly regulated genes, while the majority of DEGs is
significant in one comparison only.

@@

To understand which
pathways are upregulated in each of the two groups, we used a standard generation I
gene set enrichment analysis – a hypergeometric test – on the DEGs in each
group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
Gene sets with more than `r max_mod_size` or 
fewer than `r min_mod_size` genes were removed.
For each group, we have
selected only genes which are DEGs in that group, but not the other,
mimicking a naive approach for finding pathways regulated in one patient
group only.
Here, a similar picture
emerged. Just
`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. Again,
both the Venn diagram (Fig. 1, B) and the results of
enrichments (Fig. 1, C and D) suggest that there is a
fundamental difference between the groups, and that the groups have little
in common in their response to the virus.

Importantly, the different GO terms enriched in the two groups were related
to infectious diseases, and may tempt to speculate about the underlying
biological differences between these two groups (e.g.  significance of
Toll like receptor 4 pathway in G1, but not G2; and, vice versa, significance of
response to interleukin 7 in G2, but not in G1).

@@

The groups G1 and G2 were randomly sampled from the same data set.
In fact, repeated re-sampling always results in some genes being found to
be significantly different in one group, but not the other, despite the
fact that one does not expect any major differences between sets of
individuals randomly drawn from one population. Thus, the conclusions drawn
from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
inspection of genes which are DEGs in one group, but not the other
reveals the underlying statistical fallacy (Fig. 2, *A* – *D*), that is, that difference
between significant and non-significant is, in itself, not statistically
significant [@gelman2006difference]. This does not necessarily mean that
there are no differences at all between these two groups, but that lack of
significance in one group and significance in the other group does not
correctly identify differences between groups.

@@

To find genes which are differentially regulated in the two groups, the
correct statistical approach is to calculate interaction between groups
(G1, G2) and disease status (no disease vs. COVID). While it may be argued
that a test for interaction has lower power than a test for a simple
contrast, no 
genes show a significant interaction even at FDR < 0.1. In fact, this is not
surprising. The log~2~ fold changes for comparisons withing G1 and G2 are
strongly correlated (Fig. 2, *E*). For all significant genes, the Pearson correlation
coefficient is
`r format.pval(cors$sign, digits=2)`, 
while for genes
exclusively significant in G1 or G2 (genes "specific" to G1 or G2), it is
`r format.pval(cors$g1, digits=2)` 
and
`r format.pval(cors$g2, digits=2)`, 
respectively. Thus, genes which are significant in one, but not in the
other comparison often have have similar log~2~ fold changes (e.g. Fig.
2, *A*, *C* and *D*).

@@

It is worth noting that in the gene set enrichment analysis of the genes
"specific" for a given comparison – i.e., genes which are significant in
that comparison, but not significant in others – we have observed a number
of terms associated with immune response. It is a crucial point of this
manuscript to note that the spurious enrichments not only show significant
p-values, but also that the terms or pathways which appear in them are
relevant to the research hypothesis being tested. Below, we will show why
these terms (rather than random terms which have no obvious relevance for
an infectious disease) appear in the results.

To understand how gene set enrichment actually gives significant
results in such randomly generated gene sets, despite absence of genes with
significant interaction, it is first necessary to consider the definition
of a differentially expressed gene in this context. More often than not,
DEGs are defined by a threshold in p-value adjusted for multiple testing,
possibly combined with a threshold in log~2~ fold change. The commonly used
Benjamini-Hochberg procedure [@benjamini1995controlling] ensures that among
genes for which the adjusted p~adj~ < 0.05 there are at most 5% false
positives irrespective of the sample size.

This way, we can exert control over the false positive rate (FPR, type I
errors), keeping it at a relatively low level. However, we do not control
the false negative rate (FNR, type II errors).  In a powerful statistical
test (such as a t-test), the test power in a typical application will
rarely achieve more than 80%. For example, even for large
effects (Cohen's d > 0.8) and type I error rate of 0.05, a t-test only achieves
80% power with at least 
`r floor(pwr.t.test(power=.8, d=0.8)$n)`
samples per group. For small effects (Cohen's d > 0.2), the required number
of samples is at least
`r floor(pwr.t.test(power=.8, d=0.2)$n)`
per group.
Even assuming a test power of 80%, the FNR is 20%. Clearly, false negatives (FNs) occur
at much higher rates than false positives (FPs). In case of high throughput
data sets, where the FPR is controlled by Bejnamini-Hochberg procedure or a
similar technique, the FNR may be even as high as 80% [@white2019beyond].

@@

These FNs will occur
at a much higher rate within the sets of DEGs defined by the
non-overlaping areas of the VDs, that is DEGs considered to be "specific"
for one group or other in a naive approach.
To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
were drawn (Fig. 3), comparing the 
`r sum(covar$disease == "no") `
healthy controls to 
`r sum(covar$disease == "SC2") `
COVID-19 patients. Of the
`r length(g1_spec)`
genes significant in G1, but not in G2, 
`r .n1 <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n1`
(`r sprintf("%.0f%%", .n1/length(g1_spec) * 100)`) 
are significant in the full data set; 
of the
`r length(g2_spec)`
genes significant in G2, but not in G1, 
`r .n2 <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n2`
(`r sprintf("%.0f%%", .n2/length(g2_spec) * 100)`) 
are significant in the full data set.
Given that G1 and G2 were sampled from the total population, and since the
FDR was set to `r padj_thr`, we do not expect more than 
`r .nfp <- ceiling(padj_thr * sum(res$full$DEG)); .nfp`
FPs in the full data set, 
which implicates that at least 
`r .n1 + .n2 - .nfp` out of the `r length(g1_spec) + length(g2_spec)` 
"specific" DEGs are true positives in the full data set.
Thus, we can assume that at least a third of
the genes that appeared to be "specific" in the initial analysis were, in
fact, false negatives in one of the comparisons.

@@

In the example above, the groups have been randomly sampled from a larger
data set only once. Arguably, the observations might differ if the groups
were to be resampled. Furthermore, we have chosen a group size of 
`r n_stratum * 4` (`r n_stratum * 2` per group/treatment combination). 
Larger sample sizes are known to increase robustness of gene set enrichment
analysis, and group size of 20 has been shown to be relatively robust [@maleki2019size].
However, a smaller or larger group size might change the proportion of FNs in the
results and thus influence the results of gene set enrichment. Finally,
we have used a log~2~ fold change threshold of `r lfc_thr`, because raising
it would increase the fraction of FNs. However, selecting more biologically
only relevant genes which a substantially higher effect size may influence
the observed enrichments. On the other hand, raising the log~2~ fold change
threshold or lowering the sample size may lead to a smaller number of
defined DEGs, thus making the hypergeometric test less powerful and lead to
fewer gene set enrichment.

@@

@nieuwenhuis2011erroneous observed incorrect analyses of interactions in
about half of the analysed papers from top neuroscience journals where the
authors considered an experimental design allowing for a test for
interaction. We wanted to know if this problem is
common in transcriptomics, too. To this end, we have searched three
journals – from broad to specialized – for the occurence of the terms
"differential expression" with "venn diagrams" (Tab. 2). Next, we analysed
the selected papers from year 2020 (and years 2015-2020 in one case) to
decide whether the VD was described or referred to as showing genes
"specific" or "unique" to a particular group or whether a test for
interaction was performed. Finally, we checked whether gene set enrichment
analysis was applied to genes significant in one, but not the other
group in order to find gene sets and pathways regulated by one, but not
the other group.

@@

**Tab. 2** Results of the informal literature survey. 
We searched for papers using Google Scholar and the keywords "venn diagram"
and "differential expression".
*Journal*, journal title; *Years*, publication dates; *Total*, total number
of papers found using the search phrase; *Analysed*, total number of papers
which have been analysed for correctness; *Incorrect*, total number
(percentage) of
papers in which the Venn diagram was combined with comparing significance
to non-significance; *Enrichment*, total numebr (percentage) of papers
which combined the Venn diagram with a gene set enrichment analysis.

@@

We found 
that of the 
`r tab2$Analysed[nrow(tab2)]` analysed articles
 which used the terms "venn diagram" and
"differential expression", at least 
`r tab2$Incorrect[nrow(tab2)]`
were using Venn diagrams to compare
statistical significance with lack thereof by referring to "unique",
"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
`r tab$Enrichment[nrow(tab2)]`
coupled the VDs with some form of gene set enrichment analysis on the
set of supposedly "specific" DEGs. In summary, in at least a quarter of the
papers on differential expression in which a venn diagram was used, it was
illustrating an incorrect statistical procedure which may result in
artifactual gene set enrichments.

@@

Drawing conclusions from comparing significance with lack thereof is a
common statistical fallacy [@gelman2006difference]. Just as absence of evidence is not evidence of
absence, the failure to reject the null hypothesis does not consitute the
same level of evidence as rejecting it. However, when such an incorrect
analysis is combined with downstream functional analysis, the
resulting pathways or gene ontologies are misleadingly relevant. For
example, the identified gene sets are associated with immune answer for
research hypotheses involving an infectious disease, or cancer pathways if
the underlying research hypothesis involved cancer treatment. Such results
may appear reasonable in the given context, especially if the correct
analysis of interactions does not show any significant differences. This
effect is persistent or even exacerbated for larger sample sizes (Fig.
4).

We found that this type of incorrect analysis occurs in more than a quarter
of papers where the procedure was illustrated with a VD.  That is not to
say that VDs are not a useful tool, even in the context of transcriptomics
and gene set enrichments, if used correctly. For example, analysis of an
intersection of DEGs (i.e., by considering genes from the overlap in a VD)
is not an incorrect procedure.  Genes in the overlapping part of a VD are
significant in both (or all) comparisons, hence no comparison between
significance and non-significance is made.

@@

Visualization of interaction for individual genes is straightforward (see
for example Fig. 2, *A-D*). However, the point of VDs is to show a grand
overview of the whole analysis summarising thousands of results for the
analysed genes. As an alternative of such an overview, we suggest plotting
the log~2 fold changes in one comparison against log~2~ fold changes in the
second comparison. This allows an intuitive assessment of the differences
between the two comparisons, in especially with combination of color coding
the genes either are significant in the interaction or using the disco
score (see Fig. 2, *E* for an example).

Defining group-specific genes based on significant difference in one,
but no significant difference in another comparison is thus not only a
statistical fallacy, but when combined with gene set enrichment analysis
also one which can lead to potentially sound-looking results. This method
of obtaining specific differences between groups should therefore be
abandoned in favor of statistically correct approaches.
Furthermore, gene set enrichment analysis must never be applied
to sets of genes defined as significant in one comparison, but not the
other.

@@

This document has been written as an Rmarkdown file, containing all
statistical calculations required to replicate the findings and figures.
The markdown file, along with additional files required to recreate this
manuscript have been uploaded to [https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).

@@

A literature survey was performed using Google Scholar to
estimate the frequency of the incorrect use of Venn diagrams. We searched
for articles from 2020 mentioning the phrases "differential expression" and "venn
diagram" in three journals: Scientific Reports (2020), Nature Communications (2020) and
Science Immunology (2015-2020). For each of the papers identified, we
checked whether (i) the authors used the VD to show differentially expressed
transcripts significant in one comparison, but not another, (ii) the
authors discussed "unique", "non-overlapping" or "specific" regions of the
Venn diagram and (iii) whether this was coupled to gene set enrichment
analysis is any form. Articles which (i) focused only on the intersections
of the Venn diagrams (genes common to all groups), or (ii) which used the
Venn diagrams for a purpose other than to compare genes significant in one
groups, but not significant in other groups or (iii) for which a
clear-cut error could not be indicated past any reasonable doubt were not
considered incorrect.
The results of the literature survey (including links to papers classified
as incorrectly analysing the interaction) are included in the manuscript
sources.

>>>>

---
title: "Venn diagrams may indicate erroneous statistical reasoning in transcriptomics."
author: "January Weiner^1,†^, Benedikt Obermayer^1^ and Dieter Beule^1^"
date: "`r Sys.Date()`"
outputoff:
  html_document
output:
  word_document:
    reference_docx: templates/template.docx
outputoff2: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: true
bibliography: manuscript.bib
link-citations: true
XXcsl: templates/nature-medicine.csl
---

@@

A common application of differential expression analysis is 
finding genes that are differentially expressed upon treatment in only one
out of several groups of samples. One of the approaches is to test 
for significant difference in expression between treatment and control separately
in the two groups, and then select genes that show statistical
significance in one group only. This approach is then often combined with
a gene set enrichment analysis to find pathways and gene sets
regulated by treatment in only this group.

Here we show that this procedure is statistically incorrect and that the
interaction between treatment and group should be tested instead. Moreover,
we show that gene set enrichment analysis applied to such incorrectly
defined genes group-specific genes may result in misleading artifacts.
Due to the presence of false negatives, genes significant in one, but not
the other group are enriched in gene sets which which correspond to the
overall effect of the treatment.  Thus, the results appear related to the
problem at hand, but do not reflect the group-specific effect of a
treatment. A literature search revealed that more than a quarter of papers
which used a Venn diagram to illustrate the results of separate
differential analysis  have also applied this incorrect reasoning.

@@

Experimental designs for transcriptomic analyses frequently include more than one factor. Often, the
question asked is whether there is a difference between groups (first
factor) with respect to reaction to a particular treatment (second
factor). For example, we may ask whether there are differentially
expressed genes (DEGs) which are specific to a particular group of
patients, e.g. interferon response elicited by a virus in one group, but
absent in another group of patients. In other words, we ask whether the
difference between the control group (healthy subjects) and the treatment
group (infected patients) is different between two groups of individuals.
This "difference of differences" is known in statistics as an interaction
[@blalock1965theory]. To find out whether it is statistically significant,
an appriopriate statistical test for interaction should be employed.

However, another approach is widely spread [@nieuwenhuis2011erroneous].
Instead of testing the interaction, the effect of the treatment is tested
separately in both groups. Next, a difference between groups is inferred if the 
effect of treatment
is significant in one comparison, but not significant in the other. This
approach is not correct from statistical point of view, as "the difference
between significant and not significant is not itself statistically
significant" [@gelman2006difference]. For example, the p-value in the first
comparison may be 0.009, and in the other comparison 0.011. At an
alpha level of 0.01 the difference will be statistically significant in the
first, but not significant in the other comparison.

In transcriptomics, statistical tests are performed for thousands of
genes. As in the general case, the inference of differences between the
groups should correctly be done by testing the significance of interaction
between the group and the treatment. In practice, the differences between
treatment and control are frequently tested in the two groups separately.
This can be visualized using a Venn diagram (VD, Fig. 1) showing the
overall number of DEGs significant in both comparisons (the intersection in
the VD) or significant in only one comparison (the remaining two fields on
a VD).  The genes which are significant in only one comparison are
sometimes incorrectly considered as specific for the corresponding group.
Following this, gene set enrichment analysis may be used
in an attempt to test which pathways are specific to one, but not the other
group.

In this paper, we show that under reasonable assumptions this approach may
result in apparent enrichments even if there are no real statistically
significant differences
between the groups. To this end, we randomly split a cohort into two
groups, compared the treatment (viral infection) with controls in each of
the groups separately and then applied gene set enrichment to the sets of
genes significantly different in one, but not the other group.
Moreover, we show that the resulting gene set enrichments correspond to the
differential expression between treatment and control. Thus, the enriched
terms are relevant to the biological question at hand, yet while they do
reflect real processes linked to viral infection, they do not correspond to
the differences between the study groups. Finally, we use literature search to
show that this incorrect approach to study group-specific treatment effects
is not uncommon. In fact, while VDs are a useful visualisation tool also in
transcriptomics, in more than quarter of the papers where VDs were used,
group-specific genes were defined as significant in one, but not other
groups, and in 19% of the papers a gene set enrichment was performed.

@@

Consider two group of patients, G1 and G2 (Tab. 1). Each group contains 
`r n_stratum * 4`
individuals. In both groups, there is an equal number of healthy
individuals (labeled "Ctrl" on figures below) 
or patients infected with Sars-Cov-2 (labeled "SC2"). Our aim is to understand
the differences between G1 and G2 in the response to infection. For
example, we ask which genes or pathways are specifically upregulated by SC2
infection in G1 as compared to G2, and vice versa. In the following, we
used the data set GSE156063 [@mick2020upper] in two approaches (an
incorrect and the correct one) to arrive at opposite conclusions.

@@

First, we have performed differential gene expression analysis for each of
the groups G1 and G2 separately using standard bioinformatic tools. For each comparison, we defined
DEGs as genes for which the 
false discovery rate (FDR) was lower than 
`r padj_thr`
and absolute log~2~ fold change (LFC) was higher than
`r lfc_thr`. There were
`r sum(res$g1$DEG)` 
DEGs in the G1 group, and
`r sum(res$g2$DEG)`
in the G2 group. In total,
`r sum(res.merged$DEG.g1 & res.merged$DEG.g2)` 
DEGs were common for G1 and G2, 
`r sum(res.merged$DEG.g1 & !res.merged$DEG.g2)` 
DEGs were significant in G1 only ("specific" for G1), and 
`r sum(!res.merged$DEG.g1 & res.merged$DEG.g2)` 
were significant in G2 only (see Fig. 1, panel *A*).
A naive interpretation of these results implies that there is a substantial
difference between these two groups of individuals, as evidenced by a small
overlap in commonly regulated genes. The majority of DEGs is
significant in one comparison only.

@@

To understand which
pathways are upregulated in each of the two groups, we used a standard generation I
gene set enrichment analysis — a hypergeometric test — on the DEGs in each
group. Gene sets for the gene set enrichment analysis were taken from the Gene Ontology (GO) database.
Gene sets with more than `r max_mod_size` or 
fewer than `r min_mod_size` genes were removed.
For each group, we have
selected only genes which are DEGs in that group, but not the other,
mimicking a naive approach for finding pathways regulated in one patient
group only.
Here, a similar picture
emerged. Overall,
`r nrow(gsea_res$g1)` gene sets were significantly enriched in G1, and
`r nrow(gsea_res$g2)` gene sets were significantly enriched in G2. 
Both the Venn diagram (Fig. 1, B) and the results of
enrichments (Fig. 1, C and D) suggest that there is a
fundamental difference between the groups, and that the groups have little
in common in their response to the virus.

Importantly, the different GO terms enriched in the two groups were related
to infection, and may tempt to speculate about the underlying
biological differences between these two groups. For example, the significance of
Toll like receptor 4 pathway in G1, but not G2; and, vice versa, significance of
response to interleukin 7 in G2, but not in G1 may be considered as
evidence of altered immune response to the virus in G2 as compared to G1.

@@

However, the groups G1 and G2 were randomly sampled from the same data set.
In fact, repeated re-sampling always results in some genes being found to
be significantly different in one group, but not the other, despite the
fact that one does not expect any major differences between sets of
individuals randomly drawn from a single population. Thus, the conclusions drawn
from a Venn diagram-driven gene set enrichment analysis are artefactual. Closer
inspection of genes which are DEGs in one group, but not the other
reveals the underlying statistical fallacy (Fig. 2, *A* – *D*), that is, that difference
between significant and non-significant is, in itself, not statistically
significant [@gelman2006difference]. This does not necessarily mean that
there are no differences at all between these two groups, but that lack of
significance in one group and significance in the other group does not
correctly identify differences between groups.

@@

To find genes which are differentially regulated in the two groups, the
correct statistical approach is to calculate interaction between groups
(G1, G2) and disease status (no disease vs. COVID). While it may be argued
that a test for interaction has lower power than a test for a simple
contrast, no 
genes show a significant interaction even at FDR < 0.1. In fact, this is not
surprising. The log~2~ fold changes for comparisons withing G1 and G2 are
strongly correlated (Fig. 2, *E*). For all significant genes, the Pearson correlation
coefficient is
`r format.pval(cors$sign, digits=2)`, 
while for genes
exclusively significant in G1 or G2 (genes "specific" to G1 or G2), it is
`r format.pval(cors$g1, digits=2)` 
and
`r format.pval(cors$g2, digits=2)`, 
respectively. Thus, genes which are significant in one, but not in the
other comparison tend to have have similar log~2~ fold changes in both
groups (e.g. Fig.
2, *A*, *C* and *D*).

@@

It is worth noting that in the gene set enrichment analysis of the genes
"specific" for a given comparison – i.e., genes which are significant in
that comparison, but not significant in others – we have observed a number
of terms associated with immune response. It is a crucial point of this
manuscript to note that the spurious enrichments not only show significant
p-values, but also that the terms or pathways which appear in them are
relevant to the research hypothesis being tested. Below, we will show why
these terms (rather than random terms which have no obvious relevance to an
infectious disease) appear in the results.

To understand how significant results appear in a gene set enrichment
analysis in randomly generated groups despite absence of genes with
significant interaction, it is first necessary to consider the definition
of a differentially expressed gene in this context. More often than not,
DEGs are defined by a threshold in p-value adjusted for multiple testing,
possibly combined with a threshold in log~2~ fold change. The commonly used
Benjamini-Hochberg procedure [@benjamini1995controlling] ensures that among
genes for which FDR < 0.05 there are at most 5% false
positives irrespective of the sample size.

This way, we can exert control over the false positive rate (FPR, type I
errors), keeping it at a relatively low level. However, we do not control
the false negative rate (FNR, type II errors).  In a powerful statistical
test (such as a t-test), the test power in a typical application will
rarely achieve more than 80%. For example, even for large
effects (Cohen's d > 0.8) and type I error rate of 0.05, a t-test only achieves
80% power with at least 
`r floor(pwr.t.test(power=.8, d=0.8)$n)`
samples per group. For small effects (Cohen's d > 0.2), the required number
of samples is at least
`r floor(pwr.t.test(power=.8, d=0.2)$n)`
per group.
Even assuming a test power of 80%, the FNR is 20%. Clearly, false negatives (FNs) occur
at much higher rates than false positives (FPs). In the case of high throughput
data sets, where the FPR is controlled by Bejnamini-Hochberg procedure or a
similar technique, the FNR may be even as high as 80% [@white2019beyond].

@@

These FNs occur
at a much higher rate within the sets of DEGs defined by the
non-overlaping areas of the VDs, that is DEGs considered to be "specific"
for one group or other in a naive approach.
To illustrate this phenomenon, we have analysed the full data set from which G1 and G2
were drawn (Fig. 3), comparing the 
`r sum(covar$disease == "no") `
healthy controls to 
`r sum(covar$disease == "SC2") `
COVID-19 patients. Of the
`r length(g1_spec)`
genes significant in G1, but not in G2, 
`r .n1 <- res$full %>% filter(DEG & ENSEMBL %in% g1_spec) %>% nrow; .n1`
(`r sprintf("%.0f%%", .n1/length(g1_spec) * 100)`) 
are significant in the full data set; 
of the
`r length(g2_spec)`
genes significant in G2, but not in G1, 
`r .n2 <- res$full %>% filter(DEG & ENSEMBL %in% g2_spec) %>% nrow; .n2`
(`r sprintf("%.0f%%", .n2/length(g2_spec) * 100)`) 
are significant in the full data set.
Given that G1 and G2 were sampled from the total population, and since the
FDR was set to `r padj_thr`, we do not expect more than 
`r .nfp <- ceiling(padj_thr * sum(res$full$DEG)); .nfp`
FPs in the full data set, 
which implicates that at least 
`r .n1 + .n2 - .nfp` out of the `r length(g1_spec) + length(g2_spec)` 
"specific" DEGs are true positives in the full data set.
Thus, we can assume that at least a third of
the genes that appeared to be "specific" in the initial analysis were, in
fact, false negatives in one of the comparisons.

@@

In the example above, the groups have been randomly sampled from a larger
data set only once. Arguably, the observations might differ if the groups
were to be resampled. Furthermore, we have chosen a group size of 
`r n_stratum * 4` (`r n_stratum * 2` per group/treatment combination). 
Larger sample sizes are known to increase robustness of gene set enrichment
analysis, and group size of 20 has been shown to be relatively robust [@maleki2019size].
However, a smaller or larger group size might change the proportion of FNs in the
results and thus influence the results of gene set enrichment. Finally,
we have used a log~2~ fold change threshold of `r lfc_thr`, because raising
it would increase the fraction of FNs. However, filtering for biologically
relevant genes with a substantially higher effect size may influence
the observed enrichments. On the other hand, raising the log~2~ fold change
threshold or lowering the sample size may lead to a smaller number of
defined DEGs, thus making the hypergeometric test less powerful and lead to
fewer gene set enrichment.

@@

@nieuwenhuis2011erroneous observed incorrect analyses of interactions in
about half of the analysed papers from top neuroscience journals where the
authors considered an experimental design allowing for a test for
interaction. We wanted to know if this problem is
common in transcriptomics, too. To this end, we have searched three
journals – from broad to specialized – for the occurence of the terms
"differential expression" with "venn diagrams" (Tab. 2). Next, we analysed
the selected papers from year 2020 (and years 2015-2020 in one case) to
decide whether the VD was described or referred to as showing genes
"specific" or "unique" to a particular group or whether a test for
interaction was performed. Finally, we checked whether gene set enrichment
analysis was applied to genes significant in one, but not the other
group in order to find group-specific differences.

@@

**Tab. 2** Results of the informal literature survey. 
We searched for papers using Google Scholar and the keywords "venn diagram"
and "differential expression".
*Journal*, journal title; *Years*, publication dates; *Total*, total number
of papers found using the search phrase; *Analysed*, total number of papers
which have been analysed for correctness; *Incorrect*, total number
(percentage) of
papers in which the Venn diagram was combined with comparing significance
to non-significance; *Enrichment*, total number (percentage) of papers
which combined the Venn diagram with a gene set enrichment analysis.

@@

We found 
that of the 
`r tab2$Analysed[nrow(tab2)]` analysed articles
 which used the terms "venn diagram" and
"differential expression", at least 
`r tab2$Incorrect[nrow(tab2)]`
were using Venn diagrams to compare
statistical significance with lack thereof by referring to "unique",
"specific", "solely regulated" or "exclusive" DEGs. Out of these, at least 
`r tab$Enrichment[nrow(tab2)]`
coupled the VDs with some form of gene set enrichment analysis on the
set of supposedly "specific" DEGs. In summary, in at least a quarter of the
papers on differential expression in which a Venn diagram was used, it was
illustrating an incorrect statistical procedure which may result in
artifactual gene set enrichments.

@@

Drawing conclusions from comparing significance with lack thereof is a
common statistical fallacy [@gelman2006difference]. Just as absence of evidence is not evidence of
absence, the failure to reject the null hypothesis does not consitute the
same level of evidence as rejecting it. However, when such an incorrect
analysis is combined with downstream functional analysis, the
resulting pathways or gene ontologies are misleadingly relevant. For
example, the identified gene sets are associated with immune response for
a research hypotheses involving an infectious disease, or cancer pathways if
the underlying research hypothesis involved cancer treatment. Such results
may appear reasonable in the given context, especially if the correct
analysis of interactions does not show any significant differences. This
effect is persistent or even exacerbated for larger sample sizes (Fig.
4).

We found that this type of incorrect analysis occurs in more than a quarter
of papers where the procedure was illustrated with a VD.  That is not to
say that VDs are not a useful tool, even in the context of transcriptomics
and gene set enrichments, if used correctly. For example, gene set
enrichment analysis of an intersection of DEGs (i.e., by considering genes
from the overlap in a VD) is not an incorrect procedure.  Genes in the
overlapping part of a VD are significant in both (or all) comparisons,
hence no comparison between significance and non-significance is made.

@@

Visualization of interaction for individual genes is straightforward (see
for example Fig. 2, *A-D*). However, the point of VDs is to show a grand
overview of the whole analysis summarising thousands of results for the
analysed genes. As an alternative of such an overview, we suggest plotting
the log~2 fold changes in one comparison against log~2~ fold changes in the
second comparison. This allows an intuitive assessment of the differences
between the two comparisons, in especially in combination with color coding
the genes which either are significant in the interaction or by coloring
using the disco score (see Fig. 2, *E* for an example).

Defining group-specific genes based on significant difference in one,
but no significant difference in another comparison is thus more than only a
statistical fallacy leading to erroneous results. When combined with gene
set enrichment analysis it can lead to potentially
sound-looking, and therefore particularly misleading results. This method
of obtaining specific differences between groups should therefore be
abandoned in favor of statistically correct approaches.  Furthermore, gene
set enrichment analysis must never be applied to sets of genes defined as
significant in one comparison, but not the other.

@@

This manuscript has been written in Rmarkdown [@xie2018r].  
All statistical calculations required to replicate the findings and figures
are contained in the source Rmarkdown file.  The Rmarkdown file, along with
additional files required to recreate this manuscript as well as the
results
of literature survey have been uploaded to
[https://github.com/bihealth/manuscript_venn_diagrams](https://github.com/bihealth/manuscript_venn_diagrams).

@@

A literature survey was performed using Google Scholar to
estimate the frequency of the incorrect use of Venn diagrams. We searched
for articles including the phrases "differential expression" and "venn
diagram" in three journals: Scientific Reports (2020), Nature Communications (2020) and
Science Immunology (2015-2020). For each of the papers identified, we
checked whether (i) the authors used the VD to show differentially expressed
transcripts significant in one comparison, but not another, (ii) the
authors discussed "unique", "non-overlapping" or "specific" regions of the
Venn diagram and (iii) whether this was coupled to gene set enrichment
analysis is any form. Articles which (i) focused only on the intersections
of the Venn diagrams (genes common to all groups), or (ii) which used the
Venn diagrams for a purpose other than to compare genes significant in one
groups, but not significant in other groups or (iii) for which a
clear-cut error could not be identified past any reasonable doubt were not
considered incorrect.
The results of the literature survey (including links to papers classified
as incorrectly analysing the interaction) are included in the manuscript
sources.
